---
title: "Untitled"
output: html_document
date: '2022-07-18'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 9, message = FALSE, warning = FALSE)

library(tidyverse) 
library(lubridate)
library(viridis)
library(scales)
library(data.table)
library(janitor)
library(patchwork)
library(kableExtra)

theme_set(theme_light())

prices <- read_csv("./data/wfp_food_prices_mmr.csv") %>% 
  filter(!row_number() %in% c(1)) %>% 
  mutate_at(vars(longitude, latitude, price, usdprice),
            ~ as.numeric(.x)) %>% 
  mutate(date = ymd(date), 
         year = year(date)) %>% 
  mutate(source = "WFP")

# I wish I could do this whole process with something like
# mau <- dir("data, full.names = TRUE) %>% map_df(read_csv)
# But unfortunately, these files require some cleaning 

mau_files <- list.files("./data/mpr_se_data")

# Empty vector to store results 

mau_list <- vector("list", length(mau_files))

# The for loop for cleaning and merging 

for(i in 1:length(mau_files)) {
  
 mau_list[[i]] <- read_csv(paste0("./data/mpr_se_data/", mau_files[i])) %>%
   clean_names() %>%
   rename(commodity = x1) %>%
   # Note that you can't just specify column types since the number of 
   # columns is not consistent across the files
   mutate_at(vars(-c(commodity, category)), ~ as.numeric(.x)) %>%
   pivot_longer(-c(commodity, category), 
         names_to = "location", values_to = "price") %>% 
   mutate(filename = paste0(mau_files[i]),
          date = str_remove(filename, "MPR_SE_"),
          date = str_remove(date, ".csv"), 
          date = ym(date) + days(14), 
          location = str_to_title(location)) %>% 
   filter(!is.na(price)) %>% 
   # Converting viss into kg
   mutate(price = ifelse(str_detect(commodity, "viss"), price * 1.63293, price),
          commodity = str_replace(commodity, "viss", "kg"))
 
 mau <- bind_rows(mau_list) %>%
   rename(admin2 = location) %>% 
   left_join(prices %>% select(admin1, admin2), by = "admin2") %>% 
   mutate(admin1 = case_when(admin2 == "Hpapun" ~ "Kayin", 
                            admin2 == "Taungoo" ~ "Bago", 
                            admin2 == "Hpruso" ~ "Kayah", 
                            TRUE ~ admin1)) %>%
   distinct(commodity, admin1, admin2, price, date, filename) %>% 
   mutate(commodity = recode(commodity, 
                            "Cheaper Rice (1 kg)" = "Rice (low quality)", 
                            "Palm Oil (1 liter)" = "Oil (palm)", 
                            "Pulse (1 kg)" = "Pulses", 
                            "Onion (1 kg)" = "Onions (local)", 
                            "Chicken (1 kg)" = "Meat (chicken)"), 
         market = NA_real_, 
         year = year(date))
    
}

# show_col(hue_pal()(6)) 
# hue_pal()(6) %>% print()

```

## Introduction 
Data can be accessed from [WFP](https://data.humdata.org/dataset/wfp-food-prices-for-myanmar?force_layout=desktop) and [Mercy Corps' Market Analysis Unit (MAU)](http://themimu.info/market-analysis-unit). 

### Food prices vs. inflation

Inflation in food commodities has largely outstripped the overall pace of inflation in Myanmar as a whole. In the plot below, commodities which have been sufficiently measured (more than 1,000 observations between 2020 and the present), the bars are coloured on the number of observations where prices have increased more than 1% a month -- for reference, this is roughly equivalent to the official inflation rate of 12.63% (Central Statistics Organisation) for the whole of 2021. 

Price increases are particularly predominant in palm oil. Tomatoes were the only commodity to have not experienced overall price increases beyond Myanmar's inflation rate; however, as will be examined, its prices have undergone significant volatility. 

<br>

```{r inflation-vs-food}
prices %>% 
  filter(pricetype == "Retail") %>% 
  select(commodity, admin1, admin2, date, price, market, year) %>% 
  rbind(mau %>% select(-filename)) %>% 
  filter(year >= 2021) %>%  
  group_by(commodity) %>% 
  mutate(count = sum(!is.na(commodity))) %>% 
  ungroup() %>% 
  filter(price > 20) %>%
  group_by(date, commodity, admin2, admin1) %>% 
  summarise(price = mean(price)) %>% 
  arrange(date, commodity) %>% 
  group_by(commodity, admin2) %>% 
  mutate(mom = (price - lag(price)) / lag(price)) %>%
  filter(!is.na(mom) & mom < .5 & mom > -.5) %>% 
  summarise(mom = mean(mom, na.rm = TRUE),
            price = mean(price, na.rm = TRUE)) %>% 
  filter(!is.nan(mom)) %>%
  filter(commodity %in% c("Salt", "Tomatoes (local)", "Rice (low quality)",
                          "Pulses", "Oil (palm)", "Onions (local)")) %>%
  mutate(increase = ifelse(mom > 0.01, "yes", "no"),) %>% 
  group_by(commodity, increase) %>% 
  summarise(count = n()) %>%
  ggplot(aes(x = count, y = commodity, fill = fct_rev(increase), group = increase)) + 
  geom_col(position = position_dodge(width = .9)) + 
  scale_fill_manual(values = c("#F4511E", "#26C6DA")) + 
  labs(x = "Number of observations (aggregated at district per month)", 
       y = "", 
       fill = "Price\nincreased\n>1% per \nmonth", 
       title = "Food prices increased more than country's inflation rate between 2021/01 - 2022/05", 
       subtitle = "Only commonly surveyed commodities; price changes aggregated at district level", 
       caption = "Data sources: WFP and Mercy Corps MAU")
  
```
<br>


```{r table-commodities-mom}
prices %>% 
  filter(pricetype == "Retail") %>% 
  select(commodity, admin1, admin2, date, price, market, year) %>% 
  rbind(mau %>% select(-filename)) %>% 
  filter(year >= 2021 & price > 20) %>%  
  group_by(date, commodity, admin1, admin2) %>% 
  summarise(price = mean(price)) %>% 
  arrange(date, commodity) %>% 
  group_by(commodity, admin1, admin2) %>% 
  mutate(mom = (price - lag(price)) / lag(price)) %>%  
  filter(!is.na(mom)) %>%
  filter(commodity %in% c("Salt", "Tomatoes (local)", "Rice (low quality)", 
                          "Pulses", "Oil (palm)", "Onions (local)")) %>% 
  group_by(commodity) %>% 
  summarise(mom = round(mean(mom, na.rm = TRUE) * 100, digits = 2)) %>% 
  rename(`%month_on_month_change` = mom) %>% 
  kable(caption = "Average month-on-month increase in prices\n2021/01 - 2022/05") %>% 
  kable_classic_2("striped", position = "left", full_width = FALSE) %>% 
  footnote("Data sources: WFP and Mercy Corps MAU", 
           general_title = "")

```

<br><br><br>


### Food prices over time

Once aggregated at state level, Shan East is visibly an area of concern, having experienced the largest or near-the-largest price increases in salt, rice, tomatoes, pulses and onions. Whilst price shocks are prevalent across the entire country, the situation does not appear to be as severe in Bago.  

<br>


```{r fig.height=6}
prices %>% 
  filter(pricetype == "Retail") %>% 
  select(commodity, admin1, admin2, date, price, market, year) %>% 
  rbind(mau %>% select(-filename)) %>% 
  filter(year >= 2020) %>%  
  group_by(commodity) %>% 
  mutate(count = sum(!is.na(commodity))) %>% 
  ungroup() %>% 
  filter(count > 500 & price > 20) %>%
  group_by(date, commodity, admin1) %>% 
  summarise(price = mean(price)) %>% 
  filter(commodity %in% c("Salt", "Tomatoes (local)", "Rice (low quality)", 
                          "Pulses", "Oil (palm)", "Onions (local)")) %>% 
  group_by(commodity, admin1) %>% 
  arrange(date) %>% 
  filter(row_number() == 1 | row_number() == n()) %>%
  mutate(order = c("first", "last")) %>%
  select(-date) %>% 
  pivot_wider(names_from = order, values_from = price) %>% 
  mutate(difference = (last - first) / first, 
         # this changes the anomalous value for salt in Shan East
         difference = ifelse(difference > 2.5, 2.5, difference)) %>%
  ggplot(aes(x = difference, y = fct_rev(admin1), fill = commodity)) + 
  geom_col(position = position_dodge(width = .9)) + 
  facet_wrap(~commodity) +
  scale_fill_manual(values = c("#4686fbff", "#26A69A", "#30123bff",
                               "#e4460aff", "#e4cf5bff", "#9c179eff")) +
  scale_x_continuous(labels = percent, breaks = seq(-1, 2.5, by = .5)) +
    theme(legend.position = "none", 
        strip.text = element_text(size = 10, face = "bold"), 
        strip.background = element_rect(fill = "#212121")) + 
  guides(colour = guide_legend(nrow = 1)) +
  labs(x = "% change in prices", y = "",  
       title = "Changes in commodity prices, 2020-2022",
       subtitle = "Prices are MMK per kg, except for palm oil which is MMK per litre",
       caption = "Data sources: WFP and Mercy Corps MAU") 
  
```
<br>


just put in a datatable 

```{r}
prices %>% filter(commodity %in% c("Eggs (local)", "Tomatoes (local)")) %>%  
  filter(pricetype == "Retail") %>%  
  group_by(date, admin1, commodity) %>% 
  summarise(price = mean(price))


# Ok, this is a pretty big red flag -- eggs and tomatoes had the same price until 2021-10-15
# Who do you ask in wfp? 
```


```{r food-prices-facet, fig.height=7}
prices %>% 
  filter(pricetype == "Retail") %>% 
  select(commodity, admin1, admin2, date, price, market, year) %>% 
  rbind(mau %>% select(-filename)) %>% 
  filter(year >= 2020) %>%  
  group_by(commodity) %>% 
  mutate(count = sum(!is.na(commodity))) %>% 
  ungroup() %>% 
  filter(count > 500 & price > 20) %>%
  group_by(date, commodity, admin1) %>% 
  summarise(price = mean(price)) %>% 
  filter(commodity %in% c("Salt", "Tomatoes (local)", "Rice (low quality)",  
                          "Pulses", "Oil (palm)", "Onions (local)")) %>% 
  mutate(commodity = recode(commodity, "Eggs (local)" = "10 Eggs (local)")) %>% 
  ggplot(aes(x = date, y = price, colour = commodity, group = commodity)) +
  geom_line(size = .5) + 
  facet_wrap(~ admin1) +
  scale_x_date(breaks = "1 year", date_labels = "%Y") + 
  scale_colour_manual(values = c("#4686fbff", "#26A69A", "#30123bff", 
                                 "#e4460aff", "#e4cf5bff", "#9c179eff")) +
  scale_y_continuous(labels = comma) + 
  labs(x = "", y = "Price in MMK", colour = "", 
       title = "Average food commodity prices per state, 2020-2022",
       subtitle = "Prices are MMK per kg, except for palm oil which is MMK per litre",
       caption = "Data sources: WFP and Mercy Corps MAU") + 
  theme(legend.position = "top", 
        strip.text = element_text(size = 10, face = "bold"), 
        strip.background = element_rect(fill = "#212121")) + 
  guides(colour = guide_legend(nrow = 1))
  

```


Palm oil is most expensive in Sagaing, though all areas have experienced significant price increases. Shan East and Kayah are notable for their large and constant price shocks, with food prices spiking repeatedly. It is also apparent that palm oil, pulses and tomatoes have all experienced significant price volatility, which will be discussed in the next section. 


### Price shocks and volatility

Price volatility, represented below by the standard deviation in prices, has been most pronounced amongst palm oil, pulses and tomatoes, with tomatoes displaying extreme volatility in 2022. 

It is currently unclear whether this is due to its nature as a crop or whether this has been a general trend amongst other fruits and vegetables. Onions, which were one of the few other commodities whose prices have been measured sufficiently, has only displayed similar price fluctuations with tomatoes in Kayah and Shan North.  

<br>

```{r fig.height=6}
prices %>% 
  filter(pricetype == "Retail" & unit %in% c("KG", "L")) %>% 
  select(commodity, admin1, admin2, date, price, market, year) %>% 
  rbind(mau %>% select(-filename)) %>% 
  filter(year >= 2021) %>%  
  group_by(commodity) %>% 
  mutate(count = sum(!is.na(commodity))) %>% 
  ungroup() %>% 
  filter(price > 20) %>%
  group_by(date, commodity, admin2, admin1) %>% 
  mutate(sd = sd(price, na.rm = TRUE)) %>% 
  ungroup() %>% 
  mutate(count = ifelse(!is.na(commodity), 1, 0)) %>% 
  filter(!is.na(sd)) %>%
  filter(commodity %in% c("Salt", "Tomatoes (local)", "Rice (low quality)", 
                          "Pulses", "Oil (palm)", "Onions (local)")) %>%
  mutate(commodity = fct_reorder(commodity, sd)) %>%  
  ggplot(aes(x = sd, y = commodity, fill = commodity)) + 
  geom_boxplot(outlier.alpha = 0.5) + 
  scale_x_log10() + 
  theme(legend.position = "none", 
        strip.text = element_text(size = 10, face = "bold"), 
        strip.background = element_rect(fill = "#212121")) + 
  facet_wrap(~ year, nrow = 2) + 
  labs(x = "Standard deviation", 
       y = "", 
       title = "Price volatility amongst commonly surveyed commodities",
       caption = "Data sources: WFP and Mercy Corps MAU")

```

```{r fig.height=6}

prices %>% 
  filter(pricetype == "Retail" & unit %in% c("KG", "L")) %>% 
  select(commodity, admin1, admin2, date, price, market, year) %>% 
  rbind(mau %>% select(-filename)) %>% 
  filter(year >= 2021) %>%  
  group_by(commodity) %>% 
  mutate(count = sum(!is.na(commodity))) %>% 
  ungroup() %>% 
  filter(price > 20) %>%
  group_by(date, commodity, admin2, admin1) %>% 
  mutate(sd = sd(price, na.rm = TRUE) * 100) %>% 
  ungroup() %>% 
  mutate(count = ifelse(!is.na(commodity), 1, 0)) %>% 
  filter(!is.na(sd)) %>%
  filter(commodity %in% c("Salt", "Tomatoes (local)", "Rice (low quality)", 
                          "Pulses", "Oil (palm)", "Onions (local)")) %>%
  mutate(com_year = paste0(commodity, " - ", year)) %>%
  ggplot(aes(x = price, y = sd, colour = commodity)) + 
  geom_jitter(alpha = .025) + 
  scale_x_log10(labels = comma, 
                breaks = c(0, 100, 300, 1000, 3000, 9000)) + 
  scale_y_log10() +  
  scale_colour_manual(values = c("#F564E3", "#00BA38", "#619CFF", 
                               "#F8766D", "#B79F00", "#00BFC4")) + 
  geom_smooth(method = "loess", 
              se = FALSE, 
              size = .7) +
  facet_wrap(~ com_year) + 
  labs(x = "Price in MMK", 
       y = "Standard deviation", 
       colour = "", 
       title = "Price volatility amongst commonly surveyed commodities", 
       subtitle = "Points are observations at individual markets", 
       caption = "Data sources: WFP and Mercy Corps MAU") +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 60, hjust = 1), 
        strip.text = element_text(size = 10, face = "bold"), 
        strip.background = element_rect(fill = "#212121"))

```






